Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures

### 2018-03-20 poster/qnmlPoster.tex
first poster version ready

\mode<presentation>
{
  \usetheme{hiitposter}
}
%\graphicspath{{imgs/}}
\usepackage{amsmath,amsthm, amssymb,mathrsfs}
\usepackage{color}
\usepackage{times}
\usepackage{subfigure}
\usepackage{bm}
\usepackage{url}
\usepackage[numbers]{natbib}
\usepackage[orientation=portrait,size=a0,scale=1.4,debug]{beamerposter}
\usepackage{blkarray}        
%\usepackage[orientation=portrait,width = 91 ,height = 122 ,scale=1.39,debug]{beamerposter} 
%\setbeamerfont{itemize/enumerate body}{size={\fontsize{25}{31}}}
%\setbeamerfont{itemize/enumerate subbody}{size=\footnotesize}

@@

\definecolor{DRed}{rgb}{0.8,0.6,0.6}
\definecolor{DGreen}{rgb}{0.6,0.7,0.6}
\definecolor{BRed}{rgb}{0.7,0,0}
\definecolor{BGreen}{rgb}{0,0.7,0}

\newcommand{\myred}[1]{{\color{BRed}#1}}
\newcommand{\mygreen}[1]{{\color{BGreen}#1}}

\newcommand{\heading}[1]{\alert{\large #1}\\}
\definecolor{myPurple}{RGB}{174, 206, 195}
%\definecolor{myPurple}{RGB}{165,206,190}

\usepackage{textpos}
\usepackage{fancybox}
\setlength{\fboxsep}{12pt}
%\usepackage{mathtools}
\usepackage{empheq}

\theoremstyle{plain}
\newtheorem{assumption}{Assumption}

\definecolor{myblue}{rgb}{.85, .85, 1.0}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newlength\mytemplen
\newsavebox\mytempbox

@@

\makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\makeatletter
\newcommand\mathcircled[1]{%
  \mathpalette\@mathcircled{#1}%
}
\newcommand\@mathcircled[2]{%
  \tikz[baseline=(math.base)] \node[red,draw,circle,inner sep=1pt] (math) {$\m@th#1#2$};%
}
\makeatother

@@

\begin{block}{Background: Structure Learning of Bayesian Networks}
\begin{columns}[T]
   \begin{column}{0.3\textwidth} % first column {{{
     \heading{Bayesian Networks}
     \vspace*{12pt}
     \begin{itemize}
     \setlength\itemsep{1em}
     \item Provide a compact way to represent a joint distribution over a random vector $X = (X_1, \ldots , X_d)$. 
     %\item Here, each $X_i$ is a categorical random variable.
     \item Consist of:
     \begin{itemize}
     \item[1.] A Directed acyclic graph $G$ which encodes the dependencies between the components of $X$.
     \item[2.] Parameters $\theta = (\theta_1,\ldots,\theta_d)$, where $\theta_i$ denotes the parameters of the conditional distribution of $X_i$ given its parents $G_i$. 
     \end{itemize}
     \item Decomposition: $P(X \mid G, \theta) = \prod_{i=1}^dP(X_i\mid X_{G_i},\theta_i)$
     \end{itemize}     
   \end{column}

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){450}
   \end{column} % }}}

@@

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){450}
   \end{column} % }}}
   \begin{column}{0.3\textwidth}% third column {{{
     \heading{Scoring Functions}
     \begin{itemize}
	 \vspace*{12pt}
	 \setlength\itemsep{1em}     
     \item[BDeu] Bayesian marginal likelihood based on Dirichlet priors. Depends on a single 
     hyperparameter $\alpha >0$ called equivalent sample size. 
     \item[BIC] Maximized log-likelihood with penalty $\frac{k}{2}\log n$, where $k$ is the number of free parameters in the network. 
     \item[fNML] Factorized Normalized Maximum likelihood. Maximized log-likelihood with penalty defined via \textit{regret} functions.
     \end{itemize}

@@

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){810}
   \end{column} % }}}

@@

	\item However, for \textbf{a single categorical data vector}, regret and one dimensional $P^1_{NML}$-code can be computed.
	%\item In $P^1_{NML}$, regret depends on $N$, the length of the categorical data vector $D_i$ and $r_i$, the number of possible values for corresponding variable $X_i$. 
	%\item Justification from \textit{Minimum description length} principle.
	\item \textbf{Trick}. Recall factorization w.r.t. graph $G$:
	$$
	P(D\mid G) = \prod_{i = 1}^{d}P(D_i \mid D_{G_i}) = \prod_{i = 1}^{d}\frac{P(D_{i,G_i})}{P(D_{G_i})}
	$$
	\item Assume there are \textbf{no independencies} among $G_i$. Now, $D_{i,G_i} = $
	\vspace*{12pt}
	\end{itemize}
	\begin{minipage}{0.45\textwidth}
	\begin{align*}
	\begin{blockarray}{c|ccc}
	D_i\phantom{s} & \phantom{..}  & D_{G_i} & \phantom{..}\\
	\begin{block}{(c|ccc)}
  	1 & 1 & 1 & 1  \\
  	0 & 1 & 0 & 0 \\
  	1 & 1 &0 & 0  \\
  	1 & 1 & 1 & 1  \\
	\end{block}
	\end{blockarray} 
	\ \to \ 
	\begin{blockarray}{c}
	\phantom{...}\\
	\begin{block}{(c)}
  	0   \\
  	1  \\
  	2  \\
  	0   \\
	\end{block}
	\end{blockarray}
 	\end{align*}
	\end{minipage}\quad
	\begin{minipage}{0.45\textwidth}
	\shadowbox{
	\parbox{\textwidth}{
	\vspace*{12pt}
	\begin{itemize}
	\item We can treat $D_{i,G_i}$ and $D_{G_i}$ as values of single variables!
	\item $P(D_{i,G_i})$ and $P(D_{i,G_i})$ can be modeled using $P_{NML}^1$-code.
	\end{itemize}}}  
	\end{minipage}

@@

	\begin{itemize}
	 \item $D$ is understood as  $n$ observations on single categorical variable with $r$ possible values.
	\item Accurate approximation for $reg(n,r)$ from \cite{Szpankowski2012}:
	\begin{align*}
    reg(n, r) &\approx  n\left(\log{\alpha} + (\alpha + 2) \log{C_\alpha} - \frac{1}{C_\alpha}\right) \\ 
    &- \frac{1}{2} \log{\left(C_\alpha + \frac{2}{\alpha}\right)},
	\end{align*}
where $\alpha = \frac{r}{n}$ and $C_\alpha = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\alpha}}$.
	\end{itemize}

	\vspace*{12pt}
	\shadowbox{\parbox{0.85\textwidth}{
	\vspace*{12pt}
	\textbf{qNML score is:}
	\begin{itemize}	
	\item \textcolor{teal}{\textbf{hyper-parameter free}}
	\item \textcolor{teal}{\textbf{consistent}}
	\item \textcolor{teal}{\textbf{score equivalent}}
	\item \textcolor{teal}{\textbf{regular}}
	\item \textcolor{teal}{\textbf{it equals NML for many models.}}
	%\item qNML agrees with NML for every network composed of fully connected sub-graphs.
	\end{itemize}}}

@@

\begin{block}{Experimental Results}
  \begin{columns}[T]
    \begin{column}{0.35\textwidth}
      \heading{Structure Learning}

	  \begin{minipage}{0.45\textwidth}
	  \begin{figure}[h]

		\includegraphics[width=\columnwidth]{shd_all.pdf}
		\caption{SHD as a function of sample size.}
		\label{fig:all_shd}
		\end{figure}
	  \end{minipage}
	  \begin{minipage}{0.45\textwidth}
	  \begin{itemize}
	  \item Data generated from real world DAG structures.
	  \item Exact DP structure learning algorithm.
	  \end{itemize}
	  \begin{figure}[h]
		\includegraphics[width=\columnwidth]{shd_rank_all.pdf}
		\caption{Average ranks for each score.}
		\label{fig:all_shd}
		\end{figure}
	  \end{minipage}

	\vspace*{12pt}
	%\textbf{qNML seems overall a safe choice in structure learning, especially with moderate and large sample sizes!} 
    \end{column}
   \begin{column}{0.005\textwidth}\linethickness{0.3ex}
      \color{myPurple} \line(0,1){590}
   \end{column} % }}}

    \begin{column}{0.23\textwidth}
    \heading{Prediction}

@@

    \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){590}
   \end{column} % }}}
    \begin{column}{0.3\textwidth}
    \heading{Conclusions}

    \end{column}
  \end{columns}
\end{block}

>>>>

\mode<presentation>
{
  \usetheme{hiitposter}
}

\usepackage{amsmath,amsthm, amssymb,mathrsfs}
\usepackage{color}
\usepackage{times}
\usepackage{subfigure}
\usepackage{booktabs}
\usepackage{bm}
\usepackage{url}
\usepackage[numbers]{natbib}
\usepackage[orientation=portrait,size=a0,scale=1.34,debug]{beamerposter}
\usepackage{blkarray}        
%\usepackage[orientation=portrait,width = 91 ,height = 122 ,scale=1.39,debug]{beamerposter} 
%\setbeamerfont{itemize/enumerate body}{size={\fontsize{25}{31}}}
%\setbeamerfont{itemize/enumerate subbody}{size=\footnotesize}

@@

\newcommand{\heading}[1]{\alert{\large #1}\\}
\definecolor{myPurple}{RGB}{174, 206, 195}

\usepackage{fancybox}
\setlength{\fboxsep}{12pt}

\usepackage{empheq}

\definecolor{myblue}{rgb}{.85, .85, 1.0}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newlength\mytemplen
\newsavebox\mytempbox

@@

\makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{block}{Background: Structure Learning of Bayesian Networks}
\begin{columns}[T]
   \begin{column}{0.3\textwidth} % first column {{{
     \heading{Bayesian Networks}
     \vspace*{12pt}
     \begin{itemize}
     \setlength\itemsep{1em}
     \item Provide a compact way to represent a joint distribution over a random vector $X = (X_1, \ldots , X_d)$. 
     \item Consist of:
     \begin{itemize}
     \item[1.] A Directed acyclic graph $G$ which encodes the dependencies between the components of $X$.
     \item[2.] Parameters $\theta = (\theta_1,\ldots,\theta_d)$, where $\theta_i$ denotes the parameters of the conditional distribution of $X_i$ given its parents $G_i$. 
     \end{itemize}
     \item Decomposition: $P(X \mid G, \theta) = \prod_{i=1}^dP(X_i\mid X_{G_i},\theta_i)$
     \end{itemize}     
   \end{column}

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){430}
   \end{column} % }}}

@@

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){430}
   \end{column} % }}}
   \begin{column}{0.3\textwidth}% third column {{{
     \heading{Scoring Functions}
     \begin{itemize}
	 \vspace*{12pt}
	 \setlength\itemsep{1em}     
     \item[BDeu] Bayesian marginal likelihood based on Dirichlet priors. Depends on a single 
     hyperparameter $\alpha >0$ called equivalent sample size. 
     \item[BIC] Maximized log-likelihood with penalty $\frac{k}{2}\log n$, where $k$ is the number of free parameters in the network. 
     \item[fNML] Factorized Normalized Maximum Likelihood. Maximized log-likelihood with penalty defined via \textit{regret} functions.
     \end{itemize}

@@

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){820}
   \end{column} % }}}

@@

	\item However, for \textbf{a single categorical data vector}, regret and one dimensional $P^1_{NML}$-code can be computed.
	\item \textbf{Trick}. Recall factorization w.r.t. graph $G$:
	$$
	P(D\mid G) = \prod_{i = 1}^{d}P(D_i \mid D_{G_i}) = \prod_{i = 1}^{d}\frac{P(D_{i,G_i})}{P(D_{G_i})}
	$$
	\item Assume there are \textbf{no independencies} among $G_i$. Now, $D_{i,G_i} = $
	\vspace*{12pt}
	\end{itemize}
	\begin{minipage}{0.45\textwidth}
	\begin{align*}
	\begin{blockarray}{c|ccc}
	D_i\phantom{s} & \phantom{..}  & D_{G_i} & \phantom{..}\\
	\begin{block}{(c|ccc)}
  	1 & 1 & 1 & 1  \\
  	0 & 1 & 0 & 0 \\
  	1 & 1 &0 & 0  \\
  	1 & 1 & 1 & 1  \\
	\end{block}
	\end{blockarray} 
	\ \to \ 
	\begin{blockarray}{c}
	\phantom{...}\\
	\begin{block}{(c)}
  	0   \\
  	1  \\
  	2  \\
  	0   \\
	\end{block}
	\end{blockarray}
 	\end{align*}
	\end{minipage}\quad
	\begin{minipage}{0.45\textwidth}
	\shadowbox{
	\parbox{\textwidth}{
	\vspace*{12pt}
	\begin{itemize}
	\item We can treat $D_{i,G_i}$ and $D_{G_i}$ as values of single variables!
	\item $P(D_{i,G_i})$ and $P(D_{i,G_i})$ can be modeled using $P_{NML}^1$-code.
	\end{itemize}}}  
	\end{minipage}

@@

	\begin{itemize}
	 \item $D$ is understood as  $n$ observations on single categorical variable with $r$ possible values.
	\item Accurate approximation for $reg(n,r)$ from \cite{Szpankowski2012}:
	\begin{align*}
    reg(n, r) &\approx  n\left(\log{\beta} + (\beta + 2) \log{C_\beta} - \frac{1}{C_\beta}\right) \\ 
    &- \frac{1}{2} \log{\left(C_\beta + \frac{2}{\beta}\right)},
	\end{align*}
where $\beta = \frac{r}{n}$ and $C_\beta = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\beta}}$.
	\end{itemize}

	\vspace*{12pt}
	\shadowbox{\parbox{0.85\textwidth}{
	\vspace*{12pt}
	\textbf{qNML score is:}
	\begin{itemize}	
	\item {\textbf{hyper-parameter free}}
	\item {\textbf{consistent}}
	\item {\textbf{score equivalent}}
	\item {\textbf{regular}}
	\item {\textbf{it equals NML for many models.}}
	\end{itemize}}}

@@

\begin{block}{Experimental Results}
  \begin{columns}[T]
    \begin{column}{0.4\textwidth}
      \heading{Structure Learning}

      \begin{minipage}{0.3\textwidth}

      \begin{itemize}
	  \item Data generated from real world DAG structures.
	  \item $5$ networks, $d  = 6,5,11,5$ and $8$, \\ edges: $ 6,4,17,4$ and $8$. 
	  \item Networks learned using exact DP-algorithm.
	  \item We measure Structural Hamming Distance (SHD).
	  %\item Average ranks (over $5$ networks and $1000$ tests) also shown.
	  \end{itemize}

	  \end{minipage}
      \begin{minipage}{0.35\textwidth}
      	\phantom{dsaasdasdasdasd}
		\begin{itemize}
	  	\item Average ranks (over $5$ networks and $1000$ tests) also shown.
	  	\end{itemize}

	  \begin{figure}[h]
		\includegraphics[width=\columnwidth]{shd_all.pdf}
		\caption{SHD as a function of sample size.}
		\label{fig:all_shd}
		\end{figure}
	  \end{minipage}
	  \begin{minipage}{0.33\textwidth}
	  \begin{figure}[h]
		\includegraphics[width=\columnwidth]{shd_rank_all.pdf}
		\caption{Average ranks for each score.}
		\end{figure}

		\shadowbox{\parbox{0.95\textwidth}{
		qNML:
		\begin{itemize}	
		\item obtains the best ranking when $n > 300$.
		\item never has the worst average ranking.
		\end{itemize}}}

	\vspace*{12pt}
    \end{column}
   \begin{column}{0.005\textwidth}\linethickness{0.3ex}
      \color{myPurple} \line(0,1){630}
   \end{column} % }}}

    \begin{column}{0.55\textwidth}
    \heading{Prediction}

    \vspace*{12pt}
	\begin{minipage}{0.3\textwidth}
	\vspace*{12pt}
	\begin{itemize}
	\item 20 UCI data sets split to train and test sets.
	\item Used training fractions: $10\%, 20\% , \ldots, 90\%$.
	%\item Graphs learned using the exact algorithm.
	\item BIC, fNML and qNML use sequential predictive (sNML) parameters.
	\item BDeu uses Bayesian predictive parameter (BPP) values with $\alpha = 1$. 
	\item Scores are ranked from 1 to 4 according to test set log-loss.
	\end{itemize}	
	\end{minipage}
	\begin{minipage}{0.35\textwidth}	
		\centering
		\tiny{
		\textcolor{HIITgreen}{Table:} Average predictive performance rank over \\ different sample sizes for different scores. \\ \textbf{Best}, \underline{worst}.    
	    \begin{table}
		%\caption{Average predictive performance rank over different sample sizes for different model selection criteria in 20 different data sets.}
		\label{tbl:preds}
		\begin{center}
		\begin{tabular}{crrrrr}
		       Data
		    & $n$
		    & \multicolumn{1}{p{0.7cm}}{\centering BDeu}
		    & \multicolumn{1}{p{0.8cm}}{\centering BIC}
		    & \multicolumn{1}{p{0.9cm}}{\centering fNML}
		    & \multicolumn{1}{p{0.9cm}}{\centering qNML}\\
		       \midrule
		 PostOpe &     90 &              2.79 &     \textbf{1.20} &  \underline{3.06} &           2.94 \\
		    Iris &    150 &  \underline{2.82} &              2.37 &     \textbf{2.27} &           2.54 \\
		    Wine &    178 &  \underline{3.23} &     \textbf{1.88} &              2.67 &           2.22 \\
		   Glass &    214 &  \underline{3.61} &              3.09 &     \textbf{1.42} &           1.88 \\
		 Thyroid &    215 &              2.55 &  \underline{3.21} &     \textbf{1.80} &           2.44 \\
		 HeartSt &    270 &  \underline{3.12} &     \textbf{1.39} &              3.12 &           2.37 \\
		 BreastC &    286 &  \underline{3.09} &     \textbf{1.41} &              2.97 &           2.53 \\
		 HeartHu &    294 &  \underline{3.18} &     \textbf{1.66} &              2.90 &           2.27 \\
		 HeartCl &    303 &  \underline{3.46} &     \textbf{1.38} &              2.99 &           2.17 \\
		   Ecoli &    336 &              3.20 &  \underline{3.53} &     \textbf{1.24} &           2.04 \\
		   Liver &    345 &  \underline{3.17} &              2.39 &              2.69 &  \textbf{1.75} \\
		 Balance &    625 &  \underline{3.35} &              1.91 &     \textbf{1.59} &           3.16 \\
		 BcWisco &    699 &  \underline{3.06} &              2.03 &              2.89 &  \textbf{2.02} \\
		 Diabete &    768 &  \underline{2.91} &              2.70 &              2.68 &  \textbf{1.71} \\
		 TicTacT &    958 &  \underline{3.44} &              2.71 &     \textbf{1.31} &           2.53 \\
		   Yeast &   1484 &              2.60 &  \underline{3.76} &     \textbf{1.55} &           2.10 \\
		 Abalone &   4177 &              2.60 &  \underline{3.64} &     \textbf{1.04} &           2.72 \\
		 PageBlo &   5473 &              2.24 &  \underline{3.61} &     \textbf{1.31} &           2.83 \\
		   Adult &  32561 &              3.23 &  \underline{3.77} &     \textbf{1.00} &           2.00 \\
		 Shuttle &  58000 &     \textbf{1.44} &  \underline{3.78} &              1.56 &           3.22 \\
		\midrule		 
		 \textcolor{teal}{Mean} &  \phantom{} & \underline{2.95}  & 2.57 & \textbf{2.10}  &2.37 \\
		 \textcolor{teal}{SD}  &  \phantom{} & 0.49 & \underline{0.90} &0.76 & \textbf{0.43} \\
		\end{tabular}
		\end{center}
		\end{table}}	
	\end{minipage}
	\begin{minipage}{0.3\textwidth}

	\shadowbox{\parbox{0.95\textwidth}{
	\begin{itemize}
	\item fNML obtains on average the best rank with qNML as the runner-up.	
	\item qNML has the lowest standard deviation. 
	\end{itemize}}}

	We also studied the number of parameters in learned models when using $10\%$ samples for structure learning.
	\vspace*{12pt}

	\shadowbox{\parbox{0.95\textwidth}{
	\begin{itemize}
	\item qNML almost always yields more parsimonious models than fNML, but more complex than BIC.	
	\end{itemize}}}
	\end{minipage}

@@
